\documentclass[12pt]{article}

\usepackage{textcomp}

\begin{document}

\begin{center}
    \textbf{\Large Presentation Analysis}\\[0.5cm]
    \large Experimental Analysis of Sorting Algorithms\\[0.5cm]
    
    Maria Maiorescu\\
    Computer Science - West University of Timisoara\\
    \today
\end{center}

\vspace{0.5cm}

\section*{1. What Worked?}

\begin{itemize}
    \item The presentation followed a clear and logical structure, moving from theoretical concepts to practical benchmarking results.
    \item The graphs and tables helped visualize the performance differences between the sorting algorithms.
    \item The methodology was explained clearly, making it easy to understand how the experiments were conducted.
    \item The conclusions were supported by the collected data and experimental results.
\end{itemize}

\section*{2. What Did Not Work?}

\begin{itemize}
    \item The presentation was delivered online, so I could not see the audience and their reactions.
    \item Because of the online format, it was difficult to know whether some explanations required additional clarification.
    \item Some slides contained many numerical values, which may have been difficult for the audience to process quickly.
\end{itemize}

\section*{3. Did You Achieve the Operational Objectives?}

Yes. The operational objectives were achieved. The presentation successfully compared six classical sorting algorithms across different input sizes and input structures. The benchmarking methodology was presented clearly, and the results demonstrated how algorithm performance changes depending on the characteristics of the input data.

\section*{4. How Would You Change the Presentation?}

\begin{itemize}
    \item I would simplify some of the result slides and focus only on the most important numerical values.
    \item I would include more visual comparisons and fewer detailed tables.
    \item If possible, I would present the project in person to improve audience interaction and receive immediate feedback.
\end{itemize}

\section*{5. Self-Assessment}

I would assess my presentation at 90\%. The presentation was well organized, technically accurate, and supported by experimental results. However, there is still room for improvement in audience interaction and in presenting complex data in a more accessible way.

\end{document}